---
title: "Infrastructure Investment & Jobs Act Funding Allocation"
author: "Pricilla"
date: "2026-02-01"
output:
  pdf_document: default
  html_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(maps)
library(knitr)
library(kableExtra)
library(ggrepel)
```





```{r}
# 1. Define the URL for the RAW CSV file
url <- "https://raw.githubusercontent.com/prnakyazze94/Data-608/refs/heads/main/IIJA%20FUNDING%20AS%20OF%20MARCH%202023."

# 2. Read the data directly into R
Funding_data <- read.csv(url, check.names = FALSE)

# 3. View the first few rows
#head(Funding_data)
```

```{r}
# 1. Define the URL POPULATION
url2 <- "https://raw.githubusercontent.com/prnakyazze94/Data-608/refs/heads/main/Population%20by%20state.csv"

# 2. Read the CSV into the variable 'Popn_by_state'
Popn_by_state <- read.csv(url2, check.names = FALSE)

# 3. View the first few rows to confirm it loaded
#head(Popn_by_state)
```

The Infrastructure Investment & Jobs Act (IIJA) represents a historic injection of capital into the American backbone. By combining federal funding allocation data with 2020 Census population figures , we can uncover not just where the money is going, but who it is actually serving.

```{r}

# --- Step 1: Reshape the Population Data ---
Popn_Long <- Popn_by_state %>%
  select(-State) %>% 
  pivot_longer(
    cols = everything(), 
    names_to = "State_Name", 
    values_to = "Total_Population"
  ) %>%
  mutate(
    State_Upper = str_to_upper(State_Name),
    Total_Population = as.numeric(gsub(",", "", Total_Population))
  )

# --- Step 2: Prepare the Funding Data ---
Funding_Ready <- Funding_data %>%
  mutate(
    State_Upper = str_to_upper(`State Teritory or Tribal Nation`),
    
    # Fix typos in the JOIN KEY (so the data matches)
    State_Upper = case_when(
      State_Upper == "DELEWARE" ~ "DELAWARE", 
      State_Upper == "MISSISIPPI" ~ "MISSISSIPPI",
      TRUE ~ State_Upper
    )
  )

# --- Step 3: Join and Fix the Display Name ---
Final_Data <- left_join(Funding_Ready, Popn_Long, by = "State_Upper") %>%
  select(
    "State Teritory or Tribal Nation", 
    "Amount in Billions", 
    `Total Population` = Total_Population
  )

# *** FIX THE DISPLAY NAME HERE ***
# This overwrites "DELEWARE" with "DELAWARE" in the final list
Final_Data$"State Teritory or Tribal Nation"[Final_Data$"State Teritory or Tribal Nation" == "DELEWARE"] <- "DELAWARE"

# --- Step 4: Check the Result ---
#print(Final_Data)
```

```{r}


# 1. Get the standard USA map data
us_map <- map_data("state")

# 2. Prepare your data for joining
Data_For_Map <- Final_Data %>%
  mutate(region = tolower(`State Teritory or Tribal Nation`)) %>%
  select(region, `Total Population`)

# 3. Join the coordinate data with your population data
Map_Plot_Data <- left_join(us_map, Data_For_Map, by = "region")

#  CREATE STATE CENTERS FOR LABELS ---
# This aggregates the map coordinates to find the center (mean) of each state
State_Centers <- Map_Plot_Data %>%
  group_by(region) %>%
  summarize(long = mean(long), lat = mean(lat)) %>%
  # Add state abbreviations (built-in R datasets state.name and state.abb)
  mutate(State_Abbrev = state.abb[match(region, tolower(state.name))])

# 4. Create the Map Plot
ggplot() +
  
  # Layer 1: The Map Polygons
  geom_polygon(data = Map_Plot_Data, 
               aes(x = long, y = lat, group = group, fill = `Total Population`),
               color = "white", size = 0.2) +
  
  # Layer 2: The State Initials
  geom_text(data = State_Centers, 
            aes(x = long, y = lat, label = State_Abbrev),
            color = "black", size = 3, fontface = "bold", check_overlap = TRUE) +
  
  # Color Scale
  scale_fill_gradient(low = "#deebf7",    
                      high = "#1f77b4",   
                      labels = scales::comma,
                      name = "Population") +
  
  # Fix aspect ratio & clean theme
  coord_map() +
  labs(title = "US Population by State",
       subtitle = "Darker blue indicates higher population") +
  theme_void() +
  theme(legend.position = "right")
```


```{r}
#which states got the least and most funding.
#    We use 'order' on the second column (the billions column)
sorted_data_high <- Final_Data[order(Final_Data[[2]], decreasing = TRUE), ]

# 2. Sort the data by amount (Low to High)
sorted_data_low <- Final_Data[order(Final_Data[[2]], decreasing = FALSE), ]

# 3. Extract the top 10 and bottom 10
top_10_highest <- head(sorted_data_high, 10)
top_10_lowest  <- head(sorted_data_low, 10)

# 4. View the results

#print(top_10_highest)
```

```{r}

#Which 10 states have the highest population?
# Rename columns ( State, Billions, Population)
colnames(top_10_highest) <- c("State", "Billions", "Population")

# Plot
#    x = reorder(State, -Population) sorts the bars so they aren't random
ggplot(top_10_highest, aes(x = reorder(State, Population), y = Population)) +
  
  # The Bars
  geom_bar(stat = "identity", fill = "#1f77b4") +
  
  # The Population Text Labels (Inside the bar, White)
  geom_text(aes(label = format(Population, big.mark = ",", scientific = FALSE)),
            hjust = 1.1,      # 1.1 places text INSIDE the bar
            color = "white",  # White text
            fontface = "bold",
            size = 3.5) +
  
  # Flip axes for horizontal bars
  coord_flip() +
  
  # Labels
  labs(title = "10 States with the highest Populations",
      # subtitle = "White numbers indicate Total Population",
       x = NULL,
       y = NULL) +  # Removed Y label since numbers are hidden
  
  # Clean Theme + Remove Axis Numbers
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),  # Removes the numbers on the bottom (X-axis visually)
    axis.ticks.x = element_blank(), # Removes the ticks on the bottom
    panel.grid = element_blank()    # Removes the grid lines for a cleaner look
  )
```


```{r}

# Rename columns (data has 3 columns: State, Billions, Population)
colnames(top_10_lowest) <- c("State", "Billions", "Population")

# Plot
#    x = reorder(State, -Population) sorts the bars so they aren't random
ggplot(top_10_lowest, aes(x = reorder(State, -Population), y = Population)) +
  
  # The Red Bars
  geom_bar(stat = "identity", fill = "#d62728") +
  
  # The Population Text Labels (Inside the bar, White)
  geom_text(aes(label = format(Population, big.mark = ",", scientific = FALSE)),
            hjust = 1.1,      # 1.1 places text INSIDE the bar
            color = "white",  # White text
            fontface = "bold",
            size = 3.5) +
  
  # Flip axes for horizontal bars
  coord_flip() +
  
  # Labels
  labs(title = "10 States with the Lowest Populations",
      # subtitle = "White numbers indicate Total Population",
       x = NULL,
       y = NULL) +  # Removed Y label since numbers are hidden
  
  # Clean Theme + Remove Axis Numbers
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),  # Removes the numbers on the bottom (X-axis visually)
    axis.ticks.x = element_blank(), # Removes the ticks on the bottom
    panel.grid = element_blank()    # Removes the grid lines for a cleaner look
  )
```
```{r}

# Which 10 states received the most funding? 
# Top 10 list based on POPULATION
top_10_pop <- Final_Data %>%
  arrange(desc(`Total Population`)) %>%
  head(10)

# Rename columns
colnames(top_10_pop) <- c("State", "Billions", "Population")

# 3. Plot with Light Blue bars
ggplot(top_10_pop, aes(x = reorder(State, Billions), y = Billions)) +
  
  # -
  geom_bar(stat = "identity", fill = "#56B4E9") + 
  # --------------------------------------------------------
  
  geom_text(aes(label = paste0("$", Billions, " B")),
            hjust = 1.1,
            color = "white", 
            fontface = "bold", 
            size = 3.5) +
  
  coord_flip() +
  
  labs(title = "10 States that received the most funding",
       #subtitle = "White numbers indicate IIJA Funding (in Billions)",
       x = NULL,
       y = "Total Population") +
  
  scale_y_continuous(labels = scales::comma) +
  
  theme_minimal()+
  theme(
    axis.text.x = element_blank(),  # Removes the numbers on the bottom (X-axis visually)
    axis.ticks.x = element_blank(), # Removes the ticks on the bottom
    panel.grid = element_blank()    # Removes the grid lines for a cleaner look
  )
```


```{r}

# Sort Final_Data by 'Amount in Billions' (Ascending) to find the lowest
lowest_funded <- Final_Data %>%
  arrange(`Amount in Billions`) %>%
  head(10)

# Rename columns to match what the plot code expects
colnames(lowest_funded) <- c("State", "Billions", "Population")

# Prepare the data (ensure Billions is numeric)
#    We create new columns to control Position (hjust) and Color based on bar size.
plot_data <- lowest_funded %>%
  mutate(
    # Threshold: If funding is less than $0.3 Billion, put text OUTSIDE (-0.1)
    # Otherwise, put it INSIDE (1.1)
    Hjust_Val = ifelse(Billions < 0.3, -0.1, 1.1),
    
    # If text is OUTSIDE, make it Black. If INSIDE, make it White.
    Color_Val = ifelse(Billions < 0.3, "black", "white")
  )

# 2. Plot
ggplot(plot_data, aes(x = reorder(State, -Billions), y = Billions)) +
  
  # The Bars
  geom_bar(stat = "identity", fill = "#ff7f7f") +
  
  # The Labels (Using the dynamic Hjust and Color columns)
  geom_text(aes(label = paste0("$", Billions, "B"),
                hjust = Hjust_Val, 
                color = Color_Val), 
            fontface = "bold", 
            size = 3.5) +
  
  # This tells ggplot to use the actual strings "black" and "white" as colors
  scale_color_identity() +
  
  # Flip axes
  coord_flip() +
  
  # Labels
  labs(title = "10 States that received the least funding",
       x = NULL,
       y = " ") +
  
  # Clean Theme
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid = element_blank()
  )
```





FUNDS VS POPULATION FOR HIGHEST FUNDED STATES

```{r}

# 1. Filter and Reshape Top 20 Data
plot_data <- Final_Data %>%
  arrange(desc(`Amount in Billions`)) %>%
  head(20) %>%
  select(`State Teritory or Tribal Nation`, `Amount in Billions`, `Total Population`) %>%
  pivot_longer(
    cols = c(`Amount in Billions`, `Total Population`),
    names_to = "Metric",
    values_to = "Value"
  ) %>%
  # --- LOGIC TO PUT CALIFORNIA INSIDE ---
  mutate(
    # If State is California, hjust = 1.1 (Inside). Others = -0.1 (Outside)
    Hjust_Val = ifelse(`State Teritory or Tribal Nation` == "CALIFORNIA", 1.1, -0.1),
    
    # If State is California, Color = White. Others = Black
    Color_Val = ifelse(`State Teritory or Tribal Nation` == "CALIFORNIA", "white", "black")
  )

# 2. Create the Plot
ggplot(plot_data, aes(x = reorder(`State Teritory or Tribal Nation`, Value), 
                      y = Value, 
                      fill = Metric)) +
  
  # Bars
  geom_bar(stat = "identity", width = 0.7, show.legend = FALSE) +
  
  # Facet: Split panels
  facet_wrap(~Metric, scales = "free_x") +
  
  # Horizontal Bars
  coord_flip() +
  
  # Add Labels using the dynamic Hjust and Color columns
  geom_text(aes(label = ifelse(Metric == "Amount in Billions", 
                               paste0("$", round(Value, 2), "B"), 
                               format(round(Value), big.mark = ",", scientific = FALSE)),
                hjust = Hjust_Val, 
                color = Color_Val), 
            size = 3.5) +
  
  # Use the actual color names ("white", "black") from our data
  scale_color_identity() +
  
  # Bar Colors
  scale_fill_manual(values = c("Amount in Billions" = "#2ca02c", 
                               "Total Population" = "#1f77b4")) +
  
  # Titles
  labs(title = "20 highest funded States: Funding vs. Population",
       
       x = NULL,
       y = NULL) +
  
  # Theme
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold", size = 12),
    plot.margin = margin(t = 10, r = 20, b = 10, l = 10)
  )
```





```{r}



# Format the 'Top 10 Highest' table
kable(top_10_highest, 
      col.names = c("State", "Funding (Billions)", "Population"), 
      digits = 2, 
      format.args = list(big.mark = ","), 
      caption = "Table 1: 10 Highest Funded States") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE, 
                position = "left")
```



Looking at funding by funds vs population for states receiving the most funding.

America’s highly populated states receive the largest funding. California, the nation's most populated state with 39.5 million people, secured the most funding of 18.4 billion dollars in funding . It is followed closely by Texas's 14.4 billion dollars and New York at 10.1 billion dollars .

This aligns with the logic of scale, massive populations require massive infrastructure. Also these bigger populations are paying more taxes. However, the correlation isn't perfect. Florida, despite being the third most populous state with 21.5 million residents , ranks fifth in funding at 8.2 billion dollars, receiving less than Illinois's 8.4 billion dollars which has nearly 9 million fewer people . This suggests that funding formulas prioritize factors beyond raw headcount or tax contributions but legacy infrastructure needs or highway lane miles.





FUNDS VS POPULATION FOR LEAST FUNDED STATES.

Which States got allocated the least funds.

```{r}

# Filter for the 20 LOWEST Funded (Sorted Ascending)
lowest_20_data <- Final_Data %>%
  arrange(`Amount in Billions`) %>% 
  head(20)

# Reshape Data
plot_data_low <- lowest_20_data %>%
  select(`State Teritory or Tribal Nation`, `Amount in Billions`, `Total Population`) %>%
  pivot_longer(
    cols = c(`Amount in Billions`, `Total Population`),
    names_to = "Metric",
    values_to = "Value"
  ) %>%
  # Since these bars are small, we default to placing text OUTSIDE (-0.1) in BLACK
  mutate(
    Hjust_Val = -0.1,
    Color_Val = "black"
  )

# Plot
ggplot(plot_data_low, aes(x = reorder(`State Teritory or Tribal Nation`, -Value), # Sorts bars correctly
                          y = Value, 
                          fill = Metric)) +
  
  # Bars
  geom_bar(stat = "identity", width = 0.7, show.legend = FALSE) +
  
  # Facet: Split panels
  facet_wrap(~Metric, scales = "free_x") +
  
  # Horizontal Bars
  coord_flip() +
  
  # Add Labels
  geom_text(aes(label = ifelse(Metric == "Amount in Billions", 
                               paste0("$", round(Value, 3), "B"), # 3 decimals for tiny amounts
                               format(round(Value), big.mark = ",", scientific = FALSE)),
                hjust = Hjust_Val, 
                color = Color_Val), 
            size = 3.5) +
  
  # Use actual color names (black) for text
  scale_color_identity() +
  
  # --- CHANGED: LIGHT GREEN and LIGHT BLUE ---
  scale_fill_manual(values = c("Amount in Billions" = "#98df8a", 
                               "Total Population" = "#aec7e8")) +
  
  # Add extra space on the right so outside labels don't get cut off
  scale_y_continuous(expand = expansion(mult = c(0, 0.4))) +
  
  # Titles
  labs(title = "20 Lowest Funded States: Funding vs. Population",
       x = NULL,
       y = NULL) +
  
  # Clean Theme
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold", size = 12),
    plot.margin = margin(t = 10, r = 20, b = 10, l = 10)
  )
```


```{r}
# Which state got allocated the least funds.
# Format the 'Lowest Funded' table
kable(top_10_lowest, 
      col.names = c("State", "Funding (Billions)", "Population"), 
      digits = 4,   # Higher precision for small numbers like 0.0686
      format.args = list(big.mark = ","), 
      caption = "Table 2: 10 Lowest Funded States") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE, 
                position = "left")
```









Small States, Heavy Wallets


While they appear at the bottom of the total funding list, America's smallest states are well funded according to ratios.

Wyoming, the least populated state in the nation approximately 577,000 residents, received 1.6 billion dollars . To put that in perspective, while California has roughly 68 times the population of Wyoming, it only received about 11 times the funding. This indicates a significant small state advantage in per capita investment, likely driven by the high fixed costs of maintaining roads and bridges across vast, sparsely populated rural areas. Similarly, Vermont and Alaska secured 1 billion dollars and 3.7 billion dollars respectively, ensuring their smaller populations are not left behind.






The Funding Floor
At the very bottom of the allocation chart are the U.S. territories, which face a vastly different reality. American Samoa received the lowest allocation at roughly 68.6 million or $0.0686 billion dollars , followed closely by the Northern Mariana Islands 83.3 million dollars .

While their populations are small, the gap between the lowest funded states and these territories is stark. For instance, Delaware the lowest funded state at 792 million dollars still received nearly 11 times more funding than American Samoa.




```{r}

# Create the plot using your full dataset (Final_Data)
ggplot(Final_Data, aes(x = `Total Population`, y = `Amount in Billions`)) +
  
  # The Dots
  geom_point(color = "#1f77b4", size = 3, alpha = 0.7) +
  
  # The Trend Line (Linear Model)
  # This shows where funding "should" be if it matched population perfectly
  geom_smooth(method = "lm", se = FALSE, color = "red", linetype = "dashed") +
  
  # Add Labels for interesting outliers (Top 5 populated + Top 5 funded + Wyoming)
  geom_text_repel(aes(label = `State Teritory or Tribal Nation`), 
                  max.overlaps = 10) +
  
  # Labels
  labs(title = "Funding vs. Population",
       subtitle = "States above the red line get more funding than their population suggests",
       x = "Total Population",
       y = "Funding (Billions)") +
  
  # Format Axis Numbers with Commas
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::dollar_format(suffix = "B")) +
  
  # Clean Theme
  theme_minimal()
```



States above the red line get more funding than their population suggests while those below get less compared to population.



The Verdict


The data tells a story of two Americas. For the densely populated states like California and Texas, the IIJA is a massive dollar value investment scaling with their big populations. But for rural states like Wyoming and Alaska, it represents a strategic, high value investment that acknowledges the high cost of connecting remote citizens.The distribution of funds under the IIJA does not correlate strictly with population magnitude. Instead, the data suggests a needs based allocation model, prioritizing states with significant infrastructure deficiencies such as highway networks or aging bridges over simple per capita metrics.


Resources
U.S. Census Bureau Website Census.gov  #https://data.census.gov/table/DECENNIALCD1182020.P1?q=populations+by+state

