---
title: "Infrastructure Investment & Jobs Act Funding Allocation 2"
author: "Pricilla"
date: "2026-02-06"
output: pdf_document
---

---
title: "Infrastructure Investment & Jobs Act Funding Allocation"
author: "Pricilla"
date: "2026-02-01"
output:
  pdf_document: default
  html_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(maps)
library(knitr)
library(kableExtra)
library(ggrepel)
library(ggplot2)
library(dplyr)
library(maps)
```



```{r}
# 1. Define the URL for the RAW CSV file
url <- "https://raw.githubusercontent.com/prnakyazze94/Data-608/refs/heads/main/IIJA%20FUNDING%20AS%20OF%20MARCH%202023."

# 2. Read the data directly into R
Funding_data <- read.csv(url, check.names = FALSE)

# 3. View the first few rows
#head(Funding_data)
```

```{r}
# 1. Define the URL POPULATION
url2 <- "https://raw.githubusercontent.com/prnakyazze94/Data-608/refs/heads/main/Population%20by%20state.csv"

# 2. Read the CSV into the variable 'Popn_by_state'
Popn_by_state <- read.csv(url2, check.names = FALSE)

# 3. View the first few rows to confirm it loaded
#head(Popn_by_state)
```


```{r}
# 1. Define the URL
url <- "https://raw.githubusercontent.com/prnakyazze94/Data-608/refs/heads/main/2020%20election%20results"

# 2. Read the data
#    check.names = FALSE keeps column names like "Popular Vote" with spaces intact if they exist
election_data <- read.csv(url, check.names = FALSE)

# 3. View the first few rows to verify
#head(election_data)
```


The Infrastructure Investment & Jobs Act (IIJA) represents a historic injection of capital into the American backbone. By combining federal funding allocation data with 2020 Census population figures , we can uncover not just where the money is going, but who it is actually serving.


```{r}

# --- Step 1: Reshape the Population Data ---
Popn_Long <- Popn_by_state %>%
  select(-State) %>% 
  pivot_longer(
    cols = everything(), 
    names_to = "State_Name", 
    values_to = "Total_Population"
  ) %>%
  mutate(
    State_Upper = str_to_upper(State_Name),
    Total_Population = as.numeric(gsub(",", "", Total_Population))
  )

# --- Step 2: Prepare the Funding Data ---
Funding_Ready <- Funding_data %>%
  mutate(
    State_Upper = str_to_upper(`State Teritory or Tribal Nation`),
    
    # Fix typos in the JOIN KEY (so the data matches)
    State_Upper = case_when(
      State_Upper == "DELEWARE" ~ "DELAWARE", 
      State_Upper == "MISSISIPPI" ~ "MISSISSIPPI",
      TRUE ~ State_Upper
    )
  )

# --- Step 3: Join and Fix the Display Name ---
Final_Data <- left_join(Funding_Ready, Popn_Long, by = "State_Upper") %>%
  select(
    "State Teritory or Tribal Nation", 
    "Amount in Billions", 
    `Total Population` = Total_Population
  )

# *** FIX THE DISPLAY NAME HERE ***
# This overwrites "DELEWARE" with "DELAWARE" in the final list
Final_Data$"State Teritory or Tribal Nation"[Final_Data$"State Teritory or Tribal Nation" == "DELEWARE"] <- "DELAWARE"

# --- Step 4: Check the Result ---
#print(Final_Data)
```



```{r}


# 1. Prepare Election Data
#    Create Full Names from Abbreviations (e.g., "AL" -> "ALABAMA")
election_prep <- election_data %>%
  mutate(
    Full_State_Name = state.name[match(State, state.abb)],
    Full_State_Name = ifelse(State == "DC", "District of Columbia", Full_State_Name),
    Full_State_Name = toupper(Full_State_Name)
  )

# 2. Prepare Final_Data
Final_Data_prep <- Final_Data %>%
  mutate(`State Teritory or Tribal Nation` = toupper(`State Teritory or Tribal Nation`))

# 3. Join and Create 'Winning_Party'
Elec_funding_data <- Final_Data_prep %>%
  inner_join(election_prep, by = c("State Teritory or Tribal Nation" = "Full_State_Name")) %>%
  mutate(
    Winning_Party = case_when(
      Electoral_Votes_Biden > Electoral_Votes_Trump ~ "Democrat",
      Electoral_Votes_Trump > Electoral_Votes_Biden ~ "Republican",
      TRUE ~ "Split"
    )
  ) %>%
  # --- FIX IS HERE ---
  # We select 'State' (the abbreviation column from election_data) directly
  select(
    State_Name = `State Teritory or Tribal Nation`, 
    `Amount in Billions`, 
    `Total Population`, 
    Abbreviation = State,  # Changed from State.y to State
    Electoral_Votes_Biden, 
    Electoral_Votes_Trump, 
    Winning_Party
  )

# 4. Check the result
#head(Elec_funding_data)
```


```{r}
#which states got the least and most funding.
#    We use 'order' on the second column (the billions column)
sorted_data_high <- Final_Data[order(Final_Data[[2]], decreasing = TRUE), ]

# 2. Sort the data by amount (Low to High)
sorted_data_low <- Final_Data[order(Final_Data[[2]], decreasing = FALSE), ]

# 3. Extract the top 10 and bottom 10
top_10_highest <- head(sorted_data_high, 10)
top_10_lowest  <- head(sorted_data_low, 20)

# 4. View the results

#print(top_10_highest)
```

**2020 Election Winner by Population Density.**


The 2020 Election Winner by Population Density shows that California and New York have the highest population for Democratic states, and Florida and Texas are the Republican states with the highest population.

```{r}


# 1. Get the standard USA map data
us_map <- map_data("state")

# 2. Prepare Data from 'Elec_funding_data'
map_data_prep <- Elec_funding_data %>%
  mutate(region = tolower(State_Name)) %>%
  select(region, `Total Population`, Winning_Party, Abbreviation)

# 3. Join with Coordinate Data
full_map_data <- left_join(us_map, map_data_prep, by = "region")

# 4. Create Centers for Labels (to put state initials in the middle)
state_centers <- full_map_data %>%
  group_by(region) %>%
  summarize(long = mean(long), 
            lat = mean(lat), 
            Abbrev = first(Abbreviation))

# 5. Create the Map Plot
ggplot() +
  
  # Map Polygons
  geom_polygon(data = full_map_data,
               aes(x = long, 
                   y = lat, 
                   group = group, 
                   fill = Winning_Party,       
                   alpha = `Total Population`), 
               color = "white", size = 0.2) +
  
  # Party Colors
  scale_fill_manual(values = c("Democrat" = "#08519c",   
                               "Republican" = "#a50f15", 
                               "Split" = "gray")) +
  
  # Shading (Legend REMOVED here)
  scale_alpha_continuous(range = c(0.4, 1), 
                         guide = "none") +  # <--- This hides the Population legend
  
  # State Initials
  geom_text(data = state_centers, 
            aes(x = long, y = lat, label = Abbrev),
            color = "black", size = 3, fontface = "bold", check_overlap = TRUE) +
  
  # Labels and Theme
  coord_map() +
  labs(title = "2020 Election Winner by Population Density",
       subtitle = "Darker shades indicate higher population",
       fill = "Winner") +
  
  theme_void() +
  theme(legend.position = "right")
```

  
  
  
Looking at funding by funds vs population for states receiving the most funding.

America’s highly populated states receive the largest funding. California, the nation's most populated state with 39.5 million people, secured the most funding of 18.4 billion dollars in funding . It is followed closely by Texas's 14.4 billion dollars and New York at 10.1 billion dollars .

This aligns with the logic of scale, massive populations require massive infrastructure. Also these bigger populations are paying more taxes. However, the correlation isn't perfect. Florida, despite being the third most populous state with 21.5 million residents , ranks fifth in funding at 8.2 billion dollars, receiving less than Illinois's 8.4 billion dollars which has nearly 9 million fewer people . This suggests that funding formulas prioritize factors beyond raw headcount or tax contributions but legacy infrastructure needs or highway lane miles.

**With High Populations, the Largest States Dominate Funding.**

When analyzing gross funding rather than per capita metrics, the states with the highest populations like California, New York, Texas and Florida dominate the distribution by receiving the largest total amounts of money in Funding. 



```{r}

# 1. Prepare Election Data (if not already done)
election_prep <- election_data %>%
  mutate(
    Full_State_Name = toupper(state.name[match(State, state.abb)]),
    Full_State_Name = ifelse(State == "DC", "DISTRICT OF COLUMBIA", Full_State_Name),
    Winning_Party = case_when(
      Electoral_Votes_Biden > Electoral_Votes_Trump ~ "Democrat",
      Electoral_Votes_Trump > Electoral_Votes_Biden ~ "Republican",
      TRUE ~ "Split"
    )
  ) %>%
  select(Full_State_Name, Winning_Party)

# 2. Filter Top 20 and Join with Party Data
top_20_party <- Final_Data %>%
  arrange(desc(`Amount in Billions`)) %>%
  head(20) %>%
  mutate(Join_Name = toupper(`State Teritory or Tribal Nation`)) %>%
  left_join(election_prep, by = c("Join_Name" = "Full_State_Name"))

# 3. Reshape and Apply Label Logic
plot_data_top <- top_20_party %>%
  mutate(Pop_For_Shading = `Total Population`) %>% 
  pivot_longer(
    cols = c(`Amount in Billions`, `Total Population`),
    names_to = "Metric",
    values_to = "Value"
  ) %>%
  mutate(
    Hjust_Val = ifelse(`State Teritory or Tribal Nation` == "CALIFORNIA", 1.1, -0.1),
    Color_Val = ifelse(`State Teritory or Tribal Nation` == "CALIFORNIA", "white", "black")
  )

# 4. Create the Plot
ggplot(plot_data_top, aes(x = reorder(`State Teritory or Tribal Nation`, Value), 
                          y = Value, 
                          fill = Winning_Party,        
                          alpha = Pop_For_Shading)) + 
  
  # Bars
  geom_bar(stat = "identity", width = 0.7) +
  
  # Facet
  facet_wrap(~Metric, scales = "free_x") +
  
  # Horizontal
  coord_flip() +
  
  # Labels
  geom_text(aes(label = ifelse(Metric == "Amount in Billions", 
                               paste0("$", round(Value, 2), "B"), 
                               format(round(Value), big.mark = ",", scientific = FALSE)),
                hjust = Hjust_Val, 
                color = Color_Val), 
            size = 3.5) +
  
  # Text Colors
  scale_color_identity() +
  
  # Party Colors
  scale_fill_manual(values = c("Democrat" = "#08519c", 
                               "Republican" = "#a50f15", 
                               "Split" = "gray50")) +
  
  # Shading (Legend Removed)
  scale_alpha_continuous(range = c(0.4, 1), 
                         guide = "none") + 
  
  # Formatting
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  
  # Titles & Theme
  labs(title = "With High Populations, the Largest States Dominate Funding",
       subtitle = "Darker bars indicate higher population.",
       x = NULL,
       y = NULL,
       fill = "2020 Winner") +
  
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold", size = 12),
    plot.margin = margin(t = 10, r = 20, b = 10, l = 10)
  )
```


**Small Populations, Big Differences in Funding.**

Wyoming and Vermont have the smallest populations among the 50 states according to the 2020 Census. While population data for U.S. territories and tribal nations is often listed separately, these locations also receive significant funding. Small Populations with big Differences in Funding can be seen by making a comparison between Wyoming that received 1.6 billions in funding with a population of 576,851 well as New Hampshire that has double the population recieved 750 million dollars.

```{r}
# 2. Filter for the 20 LOWEST Funded
lowest_20_data <- Final_Data %>%
  arrange(`Amount in Billions`) %>%
  head(20)

# 3. Join with Election Data to get Party
lowest_20_merged <- lowest_20_data %>%
  mutate(Join_Name = toupper(`State Teritory or Tribal Nation`)) %>%
  left_join(election_prep, by = c("Join_Name" = "Full_State_Name")) %>%
  # Label Territories (NA party) as "Non-Voting"
  mutate(Winning_Party = ifelse(is.na(Winning_Party), "Non-Voting", Winning_Party))

```









```{r}


# 1. Filter and Reshape (Using your existing data)
lowest_20_party <- lowest_20_merged %>% 
  arrange(`Amount in Billions`) %>%
  head(20)

plot_data_low_party <- lowest_20_party %>%
  mutate(Pop_For_Shading = `Total Population`) %>% 
  pivot_longer(
    cols = c(`Amount in Billions`, `Total Population`),
    names_to = "Metric",
    values_to = "Value"
  )

# 2. Create the Plot
ggplot(plot_data_low_party, aes(x = reorder(Join_Name, -Value), 
                                y = Value, 
                                fill = Winning_Party,      
                                alpha = Pop_For_Shading)) + 
  
  # Bars
  geom_bar(stat = "identity", width = 0.7) +
  
  # Facet
  facet_wrap(~Metric, scales = "free_x") +
  
  # Horizontal
  coord_flip() +
  
  # Add Labels
  geom_text(aes(label = ifelse(Metric == "Amount in Billions", 
                               paste0("$", round(Value, 3), "B"), 
                               format(round(Value), big.mark = ",", scientific = FALSE))),
            hjust = -0.1, 
            color = "black",
            size = 3.5,
            alpha = 1) + 
  
  # --- COLORS & SCALES ---
  
  # 1. Party Colors
  scale_fill_manual(values = c("Democrat" = "#08519c", 
                               "Republican" = "#a50f15", 
                               "Non-Voting" = "#7f7f7f",
                               "Split" = "gray50")) +
  
  # 2. Population Shading (Legend REMOVED with guide="none")
  scale_alpha_continuous(range = c(0.4, 1), 
                         guide = "none") + 
  
  # Expansion for labels
  scale_y_continuous(expand = expansion(mult = c(0, 0.45))) +
  
  # Titles
  labs(title = "Small Populations, Big Differences in Funding.",
      
       x = NULL,
       y = NULL,
       fill = "2020 Winner") +
  
  # Theme
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold", size = 12),
    plot.margin = margin(t = 10, r = 10, b = 10, l = 10),
    
    # --- LEGEND CHANGE IS HERE ---
    legend.position = "right" # Places legend on the right side, vertically centered
  )
```



**Infrastructure Funding Per Capita by State.**
  

 
Data on Infrastructure funding per capita reveals a stark contrast. Alaska, a Republican state, received approximately $5,000 per resident, giving it a significant advantage in funding basic community connectivity. In comparison, Florida, a republican densely populated state of 21 million, received less than $500 per capita. While per capita allocations for tribal nations remain unclear due to data limitations, the overall trend is evident. Under the Biden administration, Republican leaning states secured the highest funding per capita, largely due to formulas prioritizing extensive infrastructure needs in sparsely populated areas.



```{r}

# 1. Prepare Election Data
election_prep <- election_data %>%
  mutate(
    Full_State_Name = toupper(state.name[match(State, state.abb)]),
    Full_State_Name = ifelse(State == "DC", "DISTRICT OF COLUMBIA", Full_State_Name),
    Winning_Party = case_when(
      Electoral_Votes_Biden > Electoral_Votes_Trump ~ "Democrat",
      Electoral_Votes_Trump > Electoral_Votes_Biden ~ "Republican",
      TRUE ~ "Split"
    )
  ) %>%
  select(Full_State_Name, Winning_Party)

# 2. Prepare Per Capita Data for ALL States
per_capita_data <- Final_Data %>%
  # Calculate Per Capita
  mutate(Per_Capita_Funding = (`Amount in Billions` * 1000000000) / `Total Population`,
         Join_Name = toupper(`State Teritory or Tribal Nation`)) %>%
  # Join with Election Data
  left_join(election_prep, by = c("Join_Name" = "Full_State_Name")) %>%
  # Label Territories
  mutate(Winning_Party = ifelse(is.na(Winning_Party), "Non-Voting", Winning_Party))

# 3. Create the Plot
ggplot(per_capita_data, aes(x = reorder(`State Teritory or Tribal Nation`, Per_Capita_Funding), 
                            y = Per_Capita_Funding, 
                            fill = Winning_Party,       
                            alpha = `Total Population`)) + # Keep shading logic
  
  # Bars (Removed geom_text to hide funding amounts)
  geom_bar(stat = "identity", width = 0.8) +
  
  # Horizontal Bars
  coord_flip() +
  
  # --- COLORS & SCALES ---
  
  # 1. Party Colors
  scale_fill_manual(values = c("Democrat" = "#08519c",     
                               "Republican" = "#a50f15",   
                               "Non-Voting" = "#7f7f7f")) + 
  
  # 2. Population Shading (Guide = "none" removes the legend)
  scale_alpha_continuous(range = c(0.3, 1), 
                         guide = "none") + 
  
  # Titles & Theme
  labs(title = "Infrastructure Funding Per Capita by State",
       subtitle = "Color indicates 2020 Election Winner. Darker bars = Higher Population.",
       x = NULL,
       y = "Funding Per Person ($)",
       fill = "Party") + # Kept Party legend, removed Population legend
  
  theme_minimal() +
  theme(
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(size = 7), # Adjusted text size to fit 50 states
    legend.position = "right",
    plot.margin = margin(t = 10, r = 20, b = 10, l = 10)
  )
```


**Funding Rank vs. Population for Top 20 Recipients**


The Funding Rank versus Population visualization below for top 20 Recipients of funding show that 5 Republican states received the most funding per capita.

```{r}

# 1. Prepare Election Data
election_prep <- election_data %>%
  mutate(
    Full_State_Name = toupper(state.name[match(State, state.abb)]),
    Full_State_Name = ifelse(State == "DC", "DISTRICT OF COLUMBIA", Full_State_Name),
    Winning_Party = case_when(
      Electoral_Votes_Biden > Electoral_Votes_Trump ~ "Democrat",
      Electoral_Votes_Trump > Electoral_Votes_Biden ~ "Republican",
      TRUE ~ "Split"
    )
  ) %>%
  select(Full_State_Name, Winning_Party)

# 2. Calculate Ranks & Filter Top 20
rank_data <- Final_Data %>%
  mutate(
    Funding_Rank = min_rank(desc(`Amount in Billions`)),  # Rank 1 = Highest Funding
    Join_Name = toupper(`State Teritory or Tribal Nation`)
  ) %>%
  arrange(Funding_Rank) %>%
  head(20) %>% 
  left_join(election_prep, by = c("Join_Name" = "Full_State_Name"))

# 3. Reshape for Plotting
plot_data_rank <- rank_data %>%
  mutate(Pop_For_Shading = `Total Population`) %>% 
  pivot_longer(
    cols = c(Funding_Rank, `Total Population`), # Funding Rank vs Actual Population
    names_to = "Metric",
    values_to = "Value"
  ) %>%
  mutate(
    # Clean up Metric names for the strip text
    Metric = ifelse(Metric == "Funding_Rank", "Funding Rank", "Total Population")
  )

# 4. Create the Plot
ggplot(plot_data_rank, aes(x = reorder(`State Teritory or Tribal Nation`, -Value), # Sorts appropriately
                           y = Value, 
                           fill = Winning_Party,        
                           alpha = Pop_For_Shading)) + 
  
  # Bars
  geom_bar(stat = "identity", width = 0.7) +
  
  # Facet with FREE scales (Critical because Rank is 1-20 but Pop is Millions)
  facet_wrap(~Metric, scales = "free_x") +
  
  # Horizontal
  coord_flip() +
  
  # Labels: ONLY show labels for Population, hide for Rank
  geom_text(aes(label = ifelse(Metric == "Total Population", 
                               format(Value, big.mark = ",", scientific = FALSE), 
                               "")), # Empty string for Rank
            hjust = -0.1, 
            color = "black", 
            size = 3.5) +
  
  # Colors
  scale_color_identity() +
  scale_fill_manual(values = c("Democrat" = "#08519c", 
                               "Republican" = "#a50f15", 
                               "Split" = "gray50")) +
  
  # Shading (Legend Removed)
  scale_alpha_continuous(range = c(0.4, 1), guide = "none") + 
  
  # Expansion to fit long population numbers
  scale_y_continuous(expand = expansion(mult = c(0, 0.3))) +
  
  # Titles & Theme
  labs(title = "Funding Rank vs. Population for Top 20 Recipients",
       x = NULL,
       y = NULL,
       fill = "2020 Winner") +
  
  theme_minimal() +
  theme(
    axis.text.x = element_blank(), 
    axis.ticks.x = element_blank(),
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold", size = 12),
    plot.margin = margin(t = 10, r = 20, b = 10, l = 10)
  )
```



**The Verdict**



The data tells a story of two Americas. For the densely populated states like California and Texas, the IIJA is a massive dollar value investment scaling with their big populations. But for rural states like Wyoming and Alaska, it represents a strategic, high value investment that acknowledges the high cost of connecting remote citizens. The distribution of funds under the IIJA does not correlate strictly with population magnitude or political affiliation. Instead, the data suggests a needs based allocation model, prioritizing states with significant infrastructure deficiencies such as highway networks or aging bridges over simple per capita metrics regardless of Political inclining. That would be the reason why even when the democrats won the elections in 2020 Republican states received the most funding per capita.


Resources
U.S. Census Bureau Website Census.gov  #https://data.census.gov/table/DECENNIALCD1182020.P1?q=populations+by+state

https://www.fec.gov/resources/cms-content/documents/2020presgeresults.pdf
